\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-07-18 tex/paper.tex
some fixes, moved a few things around

In many fields of science such as particle physics, epidemiology,  and population
genetics, computer simulators are used to describe complex data generation processes. These simulators relate
observations $\bfx$ to the parameters $\bftheta$ of an underlying theory or mechanistic model.
In most cases, these simulators are specified as procedural implementations of forward, stochastic processes involving latent variables $\bfz$.
% and may generate either discrete or continuous data.
Rarely do these simulators admit a tractable density (or likelihood) $p(\bfx |
\bftheta)$. The prevalence and significance of this problem has motivated an
active research effort in so-called \textit{likelihood-free inference}
algorithms such as Approximate Bayesian Computation
(ABC)~\cite{beaumont2002approximate, marjoram2003markov, sisson2007sequential,
sisson2011likelihood, marin2012approximate, cranmer2015approximating}.  Often
the simulation code involves control-flow that implies the dependence of $\bfx$
on $\bfz$ is non-differentiable, making inference even more difficult and
precludes approaches such as Ref.~\citep{2016arXiv160507826G}.

@@

In parallel, with the introduction of variational
auto-encoders~\citep{DBLP:journals/corr/KingmaW13} and generative adversarial
networks~\cite{goodfellow2014generative}, there has been a vibrant research
program around implicit generative models based on neural
networks~\citep{2016arXiv161003483M}.  While these implicit generative networks
also do not admit a tractable density, they are differentiable.

@@

Generative models based on neural networks are highly parametrized and the model
parameters have no obvious interpretation. In contrast, scientific simulators
can be thought of as highly regularized generative models as they typically have
relatively few parameters and they are endowed with some level of
interpretation. In this setting, inference on the model parameters $\bftheta$ is
often of more interest than the latent variables $\bfz$.

In this note, we develop a likelihood-free inference algorithms 
non-differentiable, implicit generative mode. The first allows us 
to perform empirical Bayes through variational inference, and the 
second provides a point estimator of the parameters $\bftheta$. 
We adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives. 
The objective of both algorithms is to match marginal distribution of 
the generated data to the empirical distribution of the observations.

@@

Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the
(unknown) true distribution $p_r(\mathbf{x})$, our goal is to estimate the parameters
$\bftheta^*$ that minimize the divergence between $p_r(\mathbf{x})$ and
the implicit model $p(\mathbf{x}|\bftheta)$. That is,
\begin{equation}
    \bftheta^* = \arg \min_{\bftheta \in \bftheta} \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta)),
\end{equation}
where $\rho$ is some distance or divergence.
%~\citep{bernton2017inference}.
%For many choices of divergence $\rho$, this is asymptotically equivalent to a maximum likelihood estimate if $ \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta^*)) \to 0$.

@@

As thoroughly explored in \citep{2017arXiv170104862A}, GANs remain remarkably
difficult to train because of vanishing gradients as $d$ saturates, or because of
unreliable updates when the training procedure is relaxed. As a remedy,
Wasserstein GANs~\citep{2017arXiv170107875A} reformulate the adversarial
setup in order to minimize the Wasserstein-1 distance $W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  )$
by replacing the adversarial classifier with a 1-Lipschitz adversarial critic
$d(\cdot; \bfphi) : \mathbb{R}^d \to \mathbb{R}$, and where
\begin{align}
& W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) \nonumber \\
&= \sup_{\bfphi}  \mathbb{E}_{\tilde{\mathbf{x}}\sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}}; \bfphi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x}; \bfphi)] \nonumber \\
&\equiv \sup_{\bfphi} \mathcal{L}_W.
\end{align}
Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
for stabilizing the optimization procedure,
training $d$ and $g$ results in alternating gradient updates on $\bfphi$ and $\bftheta$ in order to respectively minimize
\begin{align}
    {\cal L}_d =\,&  \mathcal{L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
    {\cal L}_g =\,& - \mathcal{L}_W
\end{align}
%
%\begin{align}
%    {\cal L}_d =\,&  {\cal L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
%    {\cal L}_g =\,& - {\cal L}_W
%\end{align}
%\begin{align}
%    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}};\bfphi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x};\bfphi)]  \nonumber \\
%                  & + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
%    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}};\bfphi)]
%\end{align}
where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} +
(1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim
p_r(\mathbf{x})$ and $\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)$.

@@

Variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization} and evolution strategies~\citep{2011arXiv1106.4487W} are general
optimization techniques that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize,
these techniques are based on the simple fact that
\begin{equation}
    \min_{\bftheta \in {\Theta}} f(\bftheta) \leq \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] = U(\bfpsi),
\end{equation}
where $q(\bftheta|\bfpsi)$ is a proposal distribution with parameters $\bfpsi$ over input values $\bftheta$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal is flexible enough, the parameters $\bfpsi$
can be updated to place its mass arbitrarily tight around the optimum $\bftheta^* = \min_{\bftheta \in \Theta} f(\bftheta)$.

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound
$U(\bfpsi)$ is differentiable with respect to $\bfpsi$, and using the log-likelihood
trick it can be rewritten as:
\begin{align}\label{eqn:approx-grad}
    \nabla_\bfpsi U(\bfpsi) &= \nabla_\bfpsi \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] \nonumber \\
    %&= \nabla_\bfpsi \int f(\bftheta)  q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
 %   &= \int f(\bftheta) \nabla_\bfpsi q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
   % &= \int \left[ f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi) \right]  q(\bftheta|\bfpsi) d\bftheta \nonumber \\
    &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi)]
\end{align}
Effectively, this means that provided that the score function $\nabla_\bfpsi \log
q(\bftheta|\bfpsi)$ of the proposal is known and that one can evaluate
$f(\mathbf{\bftheta})$ for any $\bftheta$, then one can construct empirical
estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
$U(\bfpsi)$ with stochastic gradient descent. We will use
Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to incorporate the Natural Evolution Strategy
algorithm~\citep{2011arXiv1106.4487W} and variance reduction techniques~\citep{ranganath2014black}.
%Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the variance of the noisy estimator of the gradients may be too large to be useful in many problems. Variance reduction techniques such as Rao-Blackwellization and Control Variates can be applied here as they are in black box variational inference~\citep{ranganath2014black}.

@@

Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
variance of the noisy estimator of the gradients may be too large to be useful in many problems. Improvements are possible using variance reduction techniques such as Rao-Blackwellization,
Control Variates, and quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.
Similarly, various fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
empirical distributions are being developed~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}.
Moreover, the algorithm could be adapted to use estimates to the KL distance~\citep{perez2009estimation} without the adversarial optimization for the critic. These variants may be particularly useful for small $N$.

@@

Typically in variational inference one starts with a tractable prior
$p(\bftheta)$ and a tractable likelihood $p(\bfx |\bftheta )$, thus fixing the
posterior $p(\bftheta | \bfx )$.   After optimizing $q(\bftheta | \bfpsi)$ to
approximate the posterior, one would criticize the model through the the
posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
Effectively, we are optimizing the variational posterior $q(\bftheta | \bfpsi)$ by
directly minimizing the  divergence of the corresponding posterior predictive
distribution $\qxpsi$ and the data distribution $p_r(\bfx)$ without evaluating
the intractable likelihood $p(\bfx | \bftheta )$ or specifying a prior.
Consequently, the proposed algorithm is a form of variational inference through
model criticism~\citep{box1980sampling}.

@@

Very large penalties may
eventually make the optimization unstable, as the variance of $\nabla_\bfpsi \log q(\bftheta_m|\bfpsi)$
typically increases as the entropy of the proposal decreases.
%\kcnote{maybe, but not obvious why}

@@

As a second toy example, we consider a generator producing
5-dimensional continuous data, as originally specified in Section 4.2. of
\citep{cranmer2015approximating}. More specifically, we consider the following
generative process:
\begin{itemize}
    \item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that \\
        $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
        $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
        $z_2 \sim {\text{Mixture}}[\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \\  \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5)]$,
        $z_3 \sim {\text{Exponential}(\lambda=3)}$, and
        $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
    \item $\mathbf{x} = R  \bfz$, where $R$ is a fixed
    semi-positive definite $5 \times 5$ matrix defining a fixed projection
    of $\bfz$ into the observed space.
\end{itemize}
Again, the AVO algorithm does not have access to the density or its gradient, only samples from the generative model.
We consider observed data generated at the nominal values $\bftheta^* = (\alpha^*=1,\beta^*=-1)$.
The simulator parameters are modeled with a factored  (mean field) Gaussian
variational distribution $q(\bftheta|\bfpsi) = q(\alpha|\bfpsi) q(\beta|\bfpsi)$, where each component was
initialized with zero mean and unit variance.
Hyper-parameters are set to $M=64$, $n_\text{critic}=100$, $\lambda=0.025$, $\gamma=10$ and
Adam configured with $\alpha=0.01$, $\beta_1=0.9$ and $\beta_2=0.999$.
The architecture of the critic is the same as in the previous example.

@@

In this example, we consider observed data generated with the simplified generator of Eqn.~\ref{eq:simple_weinberg}
using $\bftheta^* = (E_\text{beam}^*=42, G_f^*=0.9)$. The simulator parameters are modeled with a
factored (mean field) Gaussian variational distribution $q(\bftheta|\bfpsi) = q(E_\text{beam}|\bfpsi)
q(G_f|\bfpsi)$, where each component is respectively initialized with mean $45$
and $1$ and variance $1$ and $0.01$. Hyper-parameters are set to $M=64$,
$n_\text{critic}=100$, $\lambda=0.0025$ and Adam configured with $\alpha=0.01$,
$\beta_1=0.9$ and $\beta_2=0.999$. As with the first example, we compare
entropy penalties $\gamma=0$ and $\gamma=5$.

@@

Adversarial Learned Inference (ALI)~\citep{dumoulin2016adversarially},
Bidirectional GANs (BiGANs)~\citep{donahue2016adversarial},
$\alpha$-GAN~\citep{rosca2017variational}, Adversarial Variational Bayes
(AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv}
algorithm of  \citep{2017arXiv170208235H}  are a recent extension to GANs that
add an inference network to the generative model.  Each of these assume a
tractable density $p(\bfx|\bftheta)$ that is differentiable with respect to
$\bftheta$, which is  not satisfied in the likelihood-free setting. Our lifting
of the non-differentiable simulator $p(\bfx|\bftheta)$ to the variational
program $q(\bfx | \bfpsi)$ provides the ability to differentiate expectations
with respect to $\bfpsi$ as in Eqn~\ref{eqn:approx-grad}; however, the density
$q(\bfx | \bfpsi)$ is still intractable. Moreover, we do not attempt to define a
recognition model  $q(\bfz, \bftheta|\bfpsi)$ as the latent space $\mathcal{Z}$
of many real-world simulators is complicated and not amenable to neural a
recognition model.
%One similarity of our work to AVB is that the ELBO is rewritten such that the intractable term $\log p(\bfz) - \log q_\bfphi(\bfz | \bfx)$ is replaced with an optimized neural network $T(x,z)$, which structurally similar to the role of the critic $d$ in the Kantorovich-Rubinstein dual description of the WGAN.

%\kcnote{does \citep{rosca2017variational} go in the above paragraph?}

@@

Perhaps the closest to our work is~\cite{bernton2017inference}, which uses
Wasserstein distance both to find point estimates $\bftheta^*$ and as a part of
a rejection sampler in an ABC-like setup (as opposed to variational inference).
They emphasize that this approach allows one to eliminate the summary statistics
typically used in ABC and calculate the Wasserstein distance explicitly, without
making use of the Kantorovich-Rubinstein duality and of a critic $d$. For high-dimensional
data, they note that this is computationally expensive and introduce an
approximation based on projection of the data onto Hilbert space-filling curves.
Their Remark 5.1 points out that~\citep{montavon2016wasserstein} proposed an
approximation of the gradient of an entropy-regularized Wasserstein distance,
which uses a similar duality. They note that ``unfortunately, it is not
applicable in the setting of purely generative models, as it involves point-wise
evaluations of the derivative of the log-likelihood.'' Thus, our contribution is
to provide gradients of an approximate MWE by taking expectations with the
variational program $q(\bfx | \bfpsi)$. This paired with the critic
Kantorovich-Rubinstein dual formulation of the Wasserstein distance
allows us to work in high dimensions without summary statistics and
optimize $\bfpsi$ with stochastic gradient descent. Additionally,
our AVO-EB procedure avoids the inefficiencies of
their Wasserstein rejection sampler incurred from using the ABC-likelihood.

@@

As reviewed in \cite{2016arXiv161003483M}, likelihood-free inference is
intimately tied to a class of algorithms that can be framed as density
estimation-by-comparison. In most cases, these inference algorithms are
formulated as an iterative two-step process where the model distribution is
first compared to the true data distribution and then updated to make it more
comparable to the latter. Relevant work in this direction includes
\citep{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood,rosca2017variational}.

@@

This work has also many connections to work on variational inference, in which the
goal is to optimize the recognition model $q(\bfz, \bftheta|\bfpsi)$ so that it
is close to the true posterior $p(\bfz, \bftheta |\mathbf{x})$. There have been efforts
to extend variational inference to intractable likelihoods; however, they
require restrictive assumptions.  In \citep{tran2017variational}, the authors
consider Variational Bayes with an Intractable Likelihood (VBIL). In that
approach ``the only requirement is that the intractable likelihood can be
estimated unbiasedly.'' In the case of simulators, they propose to use the
ABC-likelihood with an $\epsilon$ kernel. The ABC likelihood is only unbiased as
$\epsilon \to 0$,   thus this method inherits the drawbacks of the ABC-likelihood
including the choice of summary statistics and the inefficiency in evaluating
the ABC likelihood for high-dimensional data and small $\epsilon$.

>>>>

In many fields of science such as particle physics, epidemiology,  and population
genetics, computer simulators are used to describe complex data generation processes. These simulators relate
observations $\bfx$ to the parameters $\bftheta$ of an underlying theory or mechanistic model.
In most cases, these simulators are specified as procedural implementations of forward, stochastic processes involving latent variables $\bfz$.
% and may generate either discrete or continuous data.
Rarely do these simulators admit a tractable density (or likelihood) $p(\bfx |
\bftheta)$. The prevalence and significance of this problem has motivated an
active research effort in so-called \textit{likelihood-free inference}
algorithms such as Approximate Bayesian Computation
(ABC) and density estimation-by-comparison algorithms~\cite{beaumont2002approximate, marjoram2003markov, sisson2007sequential,
sisson2011likelihood, marin2012approximate, cranmer2015approximating}.
% Often
% the simulation code involves control-flow that implies the dependence of $\bfx$
% on $\bfz$ is non-differentiable, making inference even more difficult and
% precludes approaches such as Ref.~\citep{2016arXiv160507826G}.

@@

In parallel, with the introduction of variational
auto-encoders~\citep{DBLP:journals/corr/KingmaW13} and generative adversarial
networks~\cite{goodfellow2014generative}, there has been a vibrant research
program around implicit generative models based on neural
networks~\citep{2016arXiv161003483M}.  While these implicit generative networks
also do not admit a tractable density, they are differentiable.
Generative models based on neural networks are highly parametrized and the model
parameters have no obvious interpretation. In contrast, scientific simulators
can be thought of as highly regularized generative models as they typically have
relatively few parameters and they are endowed with some level of
interpretation. In this setting, inference on the model parameters $\bftheta$ is
often of more interest than the latent variables $\bfz$.

@@

In this note, we develop two likelihood-free inference algorithms
for non-differentiable, implicit generative models. The first allows us
to perform empirical Bayes through variational inference, and the
second provides a point estimator of the parameters $\bftheta$.
We adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives.
The objective of both algorithms is to match marginal distribution of
the generated data to the empirical distribution of the observations.

@@

Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the
(unknown) true distribution $p_r(\mathbf{x})$, our goal is to estimate the parameters
$\bftheta^*$ that minimize the divergence between $p_r(\mathbf{x})$ and
the implicit model $p(\mathbf{x}|\bftheta)$. That is,
\begin{equation}
    \bftheta^* = \arg \min_{\bftheta} \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta)),
\end{equation}
where $\rho$ is some distance or divergence.
%~\citep{bernton2017inference}.
%For many choices of divergence $\rho$, this is asymptotically equivalent to a maximum likelihood estimate if $ \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta^*)) \to 0$.

@@

As thoroughly explored in \citep{2017arXiv170104862A}, GANs remain remarkably
difficult to train because of vanishing gradients as $d$ saturates, or because of
unreliable updates when the training procedure is relaxed. As a remedy,
Wasserstein GANs~\citep{2017arXiv170107875A} reformulate the adversarial
setup in order to minimize the Wasserstein-1 distance $W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  )$
by replacing the adversarial classifier with a 1-Lipschitz adversarial critic
$d(\cdot; \bfphi) : \mathbb{R}^d \to \mathbb{R}$, with
\begin{align}
& W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) \nonumber \\
&= \sup_{\bfphi}  \mathbb{E}_{\tilde{\mathbf{x}}\sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}}; \bfphi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x}; \bfphi)] \nonumber \\
&\equiv \sup_{\bfphi} \mathcal{L}_W.
\end{align}
Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
for stabilizing the optimization procedure,
training $d$ and $g$ results in alternating gradient updates on $\bfphi$ and $\bftheta$ in order to respectively minimize
\begin{align}
    {\cal L}_d =\,&  \mathcal{L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
    {\cal L}_g =\,& - \mathcal{L}_W
\end{align}
%
%\begin{align}
%    {\cal L}_d =\,&  {\cal L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
%    {\cal L}_g =\,& - {\cal L}_W
%\end{align}
%\begin{align}
%    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}};\bfphi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x};\bfphi)]  \nonumber \\
%                  & + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
%    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}};\bfphi)]
%\end{align}
where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} +
(1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim
p_r(\mathbf{x})$ and $\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)$.

@@

Variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization} and evolution strategies~\citep{2011arXiv1106.4487W} are general
optimization techniques that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize,
these techniques are based on the simple fact that
\begin{equation}
    \min_{\bftheta} f(\bftheta) \leq \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] = U(\bfpsi),
\end{equation}
where $q(\bftheta|\bfpsi)$ is a proposal distribution with parameters $\bfpsi$ over input values $\bftheta$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal distribution is flexible enough, the parameters $\bfpsi$
can be updated to place its mass arbitrarily tight around the optimum $\bftheta^* = \min_{\bftheta \in \Theta} f(\bftheta)$.

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound
$U(\bfpsi)$ is differentiable with respect to $\bfpsi$, and using the log-likelihood
trick its gradient can be rewritten as:
\begin{align}\label{eqn:approx-grad}
    \nabla_\bfpsi U(\bfpsi) &= \nabla_\bfpsi \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] \nonumber \\
    %&= \nabla_\bfpsi \int f(\bftheta)  q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
 %   &= \int f(\bftheta) \nabla_\bfpsi q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
   % &= \int \left[ f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi) \right]  q(\bftheta|\bfpsi) d\bftheta \nonumber \\
    &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi)]
\end{align}
Effectively, this means that provided that the score function $\nabla_\bfpsi \log
q(\bftheta|\bfpsi)$ of the proposal is known and that one can evaluate
$f(\mathbf{\bftheta})$ for any $\bftheta$, then one can construct empirical
estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
$U(\bfpsi)$ with stochastic gradient descent. We will use
Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to make  use instead of the Natural Evolution Strategy
algorithm~\citep{2011arXiv1106.4487W} or of variance reduction techniques~\citep{ranganath2014black}.
%Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the variance of the noisy estimator of the gradients may be too large to be useful in many problems. Variance reduction techniques such as Rao-Blackwellization and Control Variates can be applied here as they are in black box variational inference~\citep{ranganath2014black}.

@@

Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
variance of the noisy estimator of the gradients may be too large to be useful in many problems. Improvements are possible using variance reduction techniques such as Rao-Blackwellization,
Control Variates, or quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.
Similarly, various fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
empirical distributions are being developed~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}.
%Moreover, the algorithm could be adapted to use estimates to the KL distance~\citep{perez2009estimation} without the adversarial optimization for the critic.
These variants may be particularly useful for small $N$.

@@

Typically in variational inference one starts with a tractable prior
$p(\bftheta)$ and a tractable likelihood $p(\bfx |\bftheta )$, thus fixing the
posterior $p(\bftheta | \bfx )$.   After optimizing $q(\bftheta | \bfpsi)$ to
approximate the posterior, one would criticize the model through the
posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
Effectively, we are optimizing the variational posterior $q(\bftheta | \bfpsi)$ by
directly minimizing the  divergence of the corresponding posterior predictive
distribution $\qxpsi$ and the data distribution $p_r(\bfx)$, without evaluating
the intractable likelihood $p(\bfx | \bftheta )$ or specifying a prior.
Consequently, the proposed algorithm is a form of variational inference through
model criticism~\citep{box1980sampling}.

@@

Finally, we note that very large penalties may
eventually make the optimization unstable, as the variance of $\nabla_\bfpsi \log q(\bftheta_m|\bfpsi)$
typically increases as the entropy of the proposal decreases.
%\kcnote{maybe, but not obvious why}

@@

As a second toy example, we consider a generator producing
5-dimensional continuous data, as originally specified in Section 4.2. of
\citep{cranmer2015approximating}. More specifically, we consider the following
generative process:
\begin{itemize}
    \item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that \\
        $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
        $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
        $z_2 \sim {\text{Mixture}}[\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \\  \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5)]$,
        $z_3 \sim {\text{Exponential}(\lambda=3)}$, and
        $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
    \item $\mathbf{x} = R  \bfz$, where $R$ is a fixed
    semi-positive definite $5 \times 5$ matrix defining a fixed projection
    of $\bfz$ into the observed space.
\end{itemize}
Again, the AVO algorithm does not have access to the density or its gradient, only samples from the generative model.
We consider observed data generated at the nominal values $\bftheta^* = (\alpha^*=1,\beta^*=-1)$.
The simulator parameters are modeled with a factored  (mean field) Gaussian
proposal distribution $q(\bftheta|\bfpsi) = q(\alpha|\bfpsi) q(\beta|\bfpsi)$, where each component was
initialized with zero mean and unit variance.
Hyper-parameters are set to $M=64$, $n_\text{critic}=100$, $\lambda=0.025$, $\gamma=10$ and
Adam configured with $\alpha=0.01$, $\beta_1=0.9$ and $\beta_2=0.999$.
The architecture of the critic is the same as in the previous example.

@@

In this example, we consider observed data generated with the simplified generator of Eqn.~\ref{eq:simple_weinberg}
using $\bftheta^* = (E_\text{beam}^*=42, G_f^*=0.9)$. The simulator parameters are modeled with a
factored (mean field) Gaussian proposal distribution $q(\bftheta|\bfpsi) = q(E_\text{beam}|\bfpsi)
q(G_f|\bfpsi)$, where each component is respectively initialized with mean $45$
and $1$ and variance $1$ and $0.01$. Hyper-parameters are set to $M=64$,
$n_\text{critic}=100$, $\lambda=0.0025$ and Adam configured with $\alpha=0.01$,
$\beta_1=0.9$ and $\beta_2=0.999$. As with the first example, we compare
entropy penalties $\gamma=0$ and $\gamma=5$.

@@

@@

Perhaps the closest to our work is~\cite{bernton2017inference}, which uses the
Wasserstein distance both to find point estimates $\bftheta^*$ and as a part of
a rejection sampler in an ABC-like setup (as opposed to variational inference).
They emphasize that this approach allows one to eliminate the summary statistics
typically used in ABC and calculate the Wasserstein distance explicitly, without
making use of the Kantorovich-Rubinstein duality and of a critic $d$. For high-dimensional
data, they note that this is computationally expensive and introduce an
approximation based on projection of the data onto Hilbert space-filling curves.
Their Remark 5.1 points out that~\citep{montavon2016wasserstein} proposed an
approximation of the gradient of an entropy-regularized Wasserstein distance,
which uses a similar duality. They note that ``unfortunately, it is not
applicable in the setting of purely generative models, as it involves point-wise
evaluations of the derivative of the log-likelihood.'' Thus, our contribution is
to provide gradients of an approximate MWE by taking expectations with the
variational program $q(\bfx | \bfpsi)$. This paired with the critic
Kantorovich-Rubinstein dual formulation of the Wasserstein distance
allows us to work in high dimensions without summary statistics and to
optimize $\bfpsi$ with stochastic gradient descent. Additionally,
our AVO-EB procedure avoids the inefficiencies of
their Wasserstein rejection sampler incurred from using the ABC-likelihood.

@@

More generally, likelihood-free
inference is intimately tied to a class of algorithms that can be framed as
density estimation-by-comparison, as reviewed in \cite{2016arXiv161003483M}. In most cases, these
inference algorithms are formulated as an iterative two-step process where the
model distribution is first compared to the true data distribution and then
updated to make it more comparable to the latter. Relevant work in this
direction includes those that rely on a classifier to estimate the discrepancy betwen the observed data and the model distributions
\citep{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood,rosca2017variational}.
Also of direct relevance in the likelihood-free setup, Hamiltonian ABC~\citep{meeds2015hamiltonian}
estimate gradients with respect
to $\bftheta$ through finite differences from multiple forward passes of the
simulator with variance reduction strategies based on controlling the source of
randomness used for the latent variable $\bfz$.

Likewise, AVO closely relates to recent extensions of GANs, such as
Adversarial Learned Inference (ALI)~\citep{dumoulin2016adversarially},
Bidirectional GANs (BiGANs)~\citep{donahue2016adversarial},
$\alpha$-GAN~\citep{rosca2017variational}, Adversarial Variational Bayes
(AVB)~\citep{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv}
algorithm of  \citep{2017arXiv170208235H}, which
add an inference network to the generative model.  Each of these assume a
tractable density $p(\bfx|\bftheta)$ that is differentiable with respect to
$\bftheta$, which is  not satisfied in the likelihood-free setting. Our lifting
of the non-differentiable simulator $p(\bfx|\bftheta)$ to the variational
program $q(\bfx | \bfpsi)$ provides the ability to differentiate expectations
with respect to $\bfpsi$ as in Eqn~\ref{eqn:approx-grad}; however, the density
$q(\bfx | \bfpsi)$ is still intractable. Moreover, we do not attempt to define a
recognition model  $q(\bfz, \bftheta|\bfpsi)$ as the latent space $\mathcal{Z}$
of many real-world simulators is complicated and not amenable to neural a
recognition model.
%One similarity of our work to AVB is that the ELBO is rewritten such that the intractable term $\log p(\bfz) - \log q_\bfphi(\bfz | \bfx)$ is replaced with an optimized neural network $T(x,z)$, which structurally similar to the role of the critic $d$ in the Kantorovich-Rubinstein dual description of the WGAN.

@@

This work has also many connections to work on variational inference, in which the
goal is to optimize the recognition model $q(\bfz, \bftheta|\bfpsi)$ so that it
is close to the true posterior $p(\bfz, \bftheta |\mathbf{x})$. There have been efforts
to extend variational inference to intractable likelihoods; however, they
require restrictive assumptions.  In \citep{tran2017variational}, the authors
consider Variational Bayes with an Intractable Likelihood (VBIL). In that
approach ``the only requirement is that the intractable likelihood can be
estimated unbiasedly.'' In the case of simulators, they propose to use the
ABC-likelihood with an $\epsilon$-kernel. The ABC likelihood is only unbiased as
$\epsilon \to 0$,   thus this method inherits the drawbacks of the ABC-likelihood
including the choice of summary statistics and the inefficiency in evaluating
the ABC likelihood for high-dimensional data and small $\epsilon$.
